The Chemistry Development Kit (CDK) \cdkversion{}: atom typing, depiction, molecular formulas, and substructure searching

### 2017-05-07 article.tex
Fix wording.
Added the BRENDA database (still using CDK fingerprints, but they're looking for a PHP alternative; closes #83)

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Use of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive,Klein2013}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ChemDes~\cite{Dong2015},
ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
and the IUPHAR/BPS Guide to Pharmacology~\cite{Southan2016} and
QSAR DataBank~\cite{Ruusmann2015} databases.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015},
prediction of organic reactions~\cite{Segler2016}, and bioactivities
of compounds~\cite{Alvarsson2016}.

@@

Substructure matching is fundamental cheminformatics operation and
plays a key role in many other functions such as fingerprint and
descriptor generation, and atom typing. Since CDK v1.2, functionality
has been added to handle the SMARTS query language. The SMARTS
language is supported well including features such as
stereochemistry, component grouping, and atom
maps (to match reaction transformations). A new \textit{Pattern} API
has been added to CDK \cdkversion{}, which simplifies finding,
filtering, and transforming search results. The API is immutable allowing a
pattern to be initialized once and then matched against several
molecules or reactions across multiple threads. During initialization
the pattern is inspected as to determine what invariants will be
needed (e.g. ring size) and only required invariants are calculated. The
internal matching algorithms provide a lazy iterator, such that the
next match is only computed when it is needed. The API handles
reactions in addition to molecules, and both can be specified as
either queries or targets.

>>>>

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Use of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive,Klein2013}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ChemDes~\cite{Dong2015},
ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
and the IUPHAR/BPS Guide to Pharmacology~\cite{Southan2016}, BRENDA~\cite{Placzek2017} and
QSAR DataBank~\cite{Ruusmann2015} databases.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015},
prediction of organic reactions~\cite{Segler2016}, and bioactivities
of compounds~\cite{Alvarsson2016}.

@@

Substructure matching is fundamental cheminformatics operation and
plays a key role in many other functions such as fingerprint and
descriptor generation, and atom typing. Since CDK v1.2, functionality
has been added to handle the SMARTS query language. The SMARTS
language is supported well including features such as
stereochemistry, component grouping, and atom
maps (to match reaction transformations). A new \textit{Pattern} API
has been added to CDK \cdkversion{}, which simplifies finding,
filtering, and transforming search results. The API is immutable allowing a
pattern to be initialized once and then matched against several
molecules or reactions across multiple threads. During initialization
the pattern is inspected so as to determine what invariants will be
needed (e.g. ring size) and only required invariants are calculated. The
internal matching algorithms provide a lazy iterator, such that the
next match is only computed when it is needed. The API handles
reactions in addition to molecules, and both can be specified as
either queries or targets.
